Diversity and evolution of centromere repeats in the maize genome

### 2014-04-19 centc.tex
DA PAPER
Almost ready to submit. will make slight changes, but have all of the
accessory figures.

To do: update pacbio table with the new numbers i have, push vincesims
to github, decide about wiping seal.

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[usenames,dvipsnames]{color}
\usepackage{natbib}

@@

Centromere repeats are found in most eukaryotes and play a critical role in kinetochore formation.  Though they exhibit considerable diversity both within and among species, little is understood about the mechanisms that drive centromere repeat evolution.  Here, we use maize as a model to investigate how a complex history involving polyploidy, fractionation, and recent domestication has impacted the diversity of the maize CentC repeat.  We first validate the existence of long tandem arrays of repeats in maize and other taxa in Zea.  We show that genetic similarity among CentC copies is highest within these arrays, suggesting that tandem duplications are the primary mechanism for the generation of new copies.  In spite of this, we see little evidence that CentC variants form distinct genetic groups, instead finding that homoplasious mutations have likely homogenized CentC diversity.  We identify Cent4, a repeat initially thought to be specific to the chromosome 4 centromere, as an LTR retrotransposon that has increased drastically in frequency through domestication.  Although the two ancestral subgenomes of maize have contributed nearly equal numbers of centromeres, our analysis shows that a vast majority of all CentC repeats derive from only one of the parental genomes.  Finally, by comparing maize with its wild progenitor teosinte, we find that the abundance of CentC has decreased through domestication.

\jri{I deleted the phrase talking about rates because we don’t really do/say anything about rate do we?}

\keywords{First keyword \and Second keyword \and More}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section*{Introduction}
\label{intro}
In spite of the rapid growth of sequenced plant genomes, plant centromeres remain poorly understood and relatively cryptic, due largely to their highly repetitive content.  Centromere repeats are highly diverse across taxa and their turnover appears to be very rapid \citep{Melters2012}. However, little is known about the genetic mechanisms that produce centromere repeat diversity.  
Domesticated maize (\emph{Zea mays} ssp. \emph{mays}) has a high quality genome assembly \citep{Schnable2009} including complete sequence of two centromeres (\jri{cite wolfgruber here?? or is schnable enough?}), and the breadth of research into maize centromeres makes it one of the best systems in which to investigate the processes governing centromere repeat evolution.
\jri{The line about old nonfunctional repeats is a good point, but we don't look for these or really say much about function.}

Maize centromeres are comprised primarily of the 156bp satellite repeat CentC and a family of retrotransposons (CRM), both of which interact with  kinetochore proteins such as CENH3 (\jri{please do citations as shown above. just put the bibtex version of the citation from bookends in the refcentc.bib file.} Wolfgruber et al 2009, Zhong et al 2002), and both repeats show considerable variation in local abundance across taxa (Albert et al 2010).  But while there is considerable effort in investigating the molecular function of maize centromere repeats (Presting Chapter and references therein? \jri{here. citing “and references therein” is sort of weak scholarship. yes, i have been guilty of it myself.}), we know comparatively little about the evolution of the sequences themselves. CRM elements are better understood, including the age and insertion preferences of different CRM families (Wolfgruber et al 2009) \jri{shouldn't we cite Anupma's PNAS paper here too?}.  In contrast,  studies \jri{what studies? reference?} have only examined the flanking sequences to CentC islands despite their known association with functional centromeres.  To date, there is no in-depth characterization of the genetic diversity of centromere repeats in the maize genome.

In this paper, we describe the patterns of diversity of centromere repeats across the maize genome.  We investigate the genetic diversity of these repeats, including whether the differential ancestry of maize centromeres (Wang and Bennetzen 2012) has led to chromosome-specific repeats as found in other species (Heslop-Harrison 1999, Hirsch et al 2009) and how genetic relatedness among individual repeats varies spatially along chromosomes and across the genome.  We then compare centromere abundance across a number of maize lines, including to its wild relatives the teosintes.  We find that CentC copies do not form genetic groups consistent with ancient whole genome duplications or chromosome specificity.  Instead, we show higher genetic similarity within clusters, potely indicating the predominance of tandem duplications in the formation of new CentC copies.  Lastly, we use low coverage sequencing and cytological analyses datshow that domesticated maize has less CentC and reli its wild relative  teosinte line

@@

We downloaded 218 previously annotated CentC sequences (Annaniev et al 1998, Nagaki et al 2003) from Genbank.  We then searched the maize genome (5b60, \url{www.maizesequence.org}) with megaBLAST (McGinnis and Madden 2004) using the 218 annotated CentCs as a reference.  We kept hits with a length of over 140bp and a minimum bit score of 100.  After meeting the bit score threshold, the longest hit was retained.  We defined CentC’s as being in tandem if the CentC’s start location was within 1000bp of the start location of another CentC.

All 12,162 CentC sequences were aligned using 7 iterations of Muscle (Edgar 2004) with default parameters.  A Jukes-Cantor distance matrix of all sequences was calculated with PHYLIP (Felsenstein 2005, \url{http://evolution.genetics.washington.edu/phylip.html}), and an unrooted neighbor joining tree was built based on the distance matrix.

We used principle coordinate analysis (PCoA) to cluster CENTC variants based on their genetic distances. Eigenvalues from the PCoA were used to determine the number of statistically significant clusters using the Tracy-Widom distribution (c.f. Patterson et al 2006).

We employed the software SpaGeDi (\url{http://ebe.ulb.ac.be/ebe/Software.html}, Hardy and Vekemans 2002) to estimate the spatial autocorrelation of sequence similarity of CENTC repeats in the completely sequenced centromeres 2 and 5.  We calculated Moran’s I statistic using Jukes-Cantor genetic distance and measures of physical distance between CENTC repeats in base pairs.  Confidence intervals for the values of I were estimated by 20,000 random permutations of the physical distances.

Statistical analyses were performed in R with the packages ape (Paradis, Claude, and Strimmer 2004) and RMTstat (Perry et al 2009).  We compared clusters to chromosome of origin and syntenic maps of maize ancient tetraploidy (Schnable et al 2011) to determine if the genetic history of maize left a footprint on CentC similarity.

@@

We mapped Illumina reads from a broad panel of Zea species \citep{Chia2012,Tenaillon2011}  to a reference consisting of the full complement of 12,162 CentC variants identified in the B73 genome.  We also used previously published whole genome chromatin immunoprecipitation (ChIP) (Wolfgruber et al 2009, Wang et al 2013) with CenH3.  Reads were mapped using  Mosaik v1.0 (\url{https://code.google.com/p/mosaik-aligner/}). We first optimized mapping parameters by relaxing mapping stringency and evaluating the number of successfully mapped reads with each combination.  Consistent with parameters from previous studies mapping reads to repetitive elements \citep{Tenaillon2011}, we required homology to remain at a minimum of 80\%.  For other non-default parameters, we permuted over many values of hash size, alignment candidate threshold,  percent of read aligning, and maximum number of hash positions per seed to find a combination that produced believable alignments.  We selected an optimum combination of parameters just below the parameters where we observed a large increase in the total number of reads aligning (Supplementary Figure X \jri{please start filling these in}).   The parameter boundary at which percentage of reads mapping increased disproportionately was identified, and parameters just below this threshold were selected.  Our final set of parameters for tandem repeats used an initial hash size of 8, an alignment candidate threshold of 15 bases, 20\% percent of mismatching bases, a minimum of 30\% overlap to the reference, and stored the top 100 hits for alignment.  After reads were mapped, we calculated the percentage of total reads hitting the given reference and multiplied this value by the relative genome size of each accession as reported in \citet{Chia2012} and \citet{Tenaillon2011}. \jri{note difference here in effect of citep and citet commands} The total number of reads mapping did not change drastically when using one random copy of CentC versus the full AGPv2 reference, suggesting that our parameters are sufficiently broad to capture genome-wide CentC.  Because library preparation has an effect on estimates of repeat abundance (see results), we only used individuals from maize HapMap v2 \citep{Chia2012} with libraries prepared using identical methods.

We used a different set of mapping parameters for long repeats such as transposable elements.  Previous studies (Schnable et al 2009) estimated that approximately 85\% of maize genome derives from transposable elements.  Using the short read libraries from Tenaillon et al (2011), we selected parameters so that approximately 85\% of the library mapped to the maize transposable element database (\url{maizetedb.org}) with a minimum homology of 80\%.  The final parameters for TEs were a hash size of 10, alignment candidate threshold of 11, 80\% homology excluding non-aligned portions of the read, and a 30\% minimum overlap.

\subsection*{CentC simulations}

\jri{needs some basic setup. we start with alignment (?) and then do what to it? what was the code written in? is the code available? if so cite github url}. Our simulation assumed that CentC had been evolving since the divergence of maize and \emph{Tripsacum} (10 million years, Ross-Ibarra et al 2009)\jri{that's not the date i cite in my paper! should be 1 million.}, a closely related genus whose centromere repeat shares a large degree of homology \citep{Melters2012}.  We assumed a constant copy number, a mutation rate of $3 x 10^{-8}$, and one generation per year.

@@

Library preparation and sequencing was performed according to the methods cited in Melters et al 2012.  Using those protocols, we sequenced one individual from \emph{mays}, \emph{mexicana}, \emph{parviglumis}, and \emph{Z. luxurians} with Pacific Biosciences (\jri{cite}) technology. Approximately 200Mbp of reads were produced from each cell, and reads with length greater than 600bp were retained for analysis of tandem CentC content using BLAST (Supplementary Table \ref{supp.pacbio} \jri{check out how tables are done and referenced with this example}).  CentC copies were considered in tandem if the read had 4 CentC copies within 300bp of each other.

@@

Flourescent in situ hybridization (FISH) was used to compare abundance of CENTC repeats in chromosomes of a maize x parviglumis F1 hybrid and a maize x Z. luxurians F1 hybrid. FISH protocols closely followed those of X and Y.
\jri{Need Kelly to fill in}

@@

We found a total of 12,162 CentC variants in the maize reference genome and the unassembled BACs.  Of these 12,162 copies, 9,896 were unique over their full length. No CentC occurred more than 10 times in the genome, and the vast majority ($>$X\% \jri{if it’s easy, this number would be nice.}) of non-unique CentC variants occurred only twice.  Of the 2,266 non-unique CentC sequences, only 3 were tandem duplicates.  Genome-wide CentC locations also show that nearly all of the 10,639 CentC copies are found in one of the 248 clusters identified on chromosomes 1-10; only 14 occurred as solo copies.  Cluster size varied from single CENTC copies to 84KB, with a mean of ~7kb (approximately 45 CentC copies). Chromosomes varied greatly in copies of CentC.   Chromosome 7, with 3,200 copies, had the most CentC, while chromosome 6 had the fewest with 32 copies.  The presence of 1,523 copies of CentC on the unassembled BACs and so few copies on chromosome 6 suggest that Centromeric regions still require further assembly.

We used long-read Pacific Biosciences sequencing to verify that most CentC is in tandem arrays. We sequenced whole genome ($\approx 0.1$X ) libraries from 4 Zea species.  In spite of the low coverage, we recovered reads containing CentC sequence from all four taxa (Supplementary Table \jri{add}).  In one 6.7kb read from the maize reference line B73, for example, we identified approximately 40 independent CentC copies in tandem, and similar arrays were seen in all four Zea species analyzed.  These results show that overall structure of the repeats has been maintained for the approximately 140,000 years since the \emph{luxurians}-\emph{mays} divergence (Hanson et al. 1996; Ross-Ibarra et al. 2009) and that a majority of CentC is found in tandem arrays (\jri{add table ref}). \jri{this last sentence is a nice addition}

\jri{seems like in this paragraph we should cite the circos figure, no?} We then identified how many large clusters of CentC were retained from each of the two parental genomes that comprise the extant maize genome.  Previous work identified the parental genome for individual chromosomal segments (Schnable et al 2011) and centromeres (Wang and Bennetzen 2012).  Because large clusters are less likely to be misassembled, we focused our analyses on the 52 clusters $>$10KB in length (Supplementary Fig Histogram \jri{add}).  While Schnable et al. (2010) do not assign subgenome status to repetitive regions, we assign clusters to a subgenome if they are flanked by two regions identified as originating from the same  subgenome.  Thirty-eight of these clusters could be assigned to subgenome 1 (out of 43 assignable); of the 17 clusters $>$20KB, 16 were unambiguously assigned to subgenome 1.  Even correcting for the genome-wide overrepresentation of subgenome 1 ( 62.5\% of assigned base pairs), these results suggest a strong inequality in the origin of large CentC clusters ( binomial test, $p<0.005$ for both 10kb and 20kb clusters). \jri{isn't this a fisher's exact?}

@@

In addition to CentC, previous studies have described a second high copy number centromere repeat specific to maize chromosome 4 (Page et al 2001).   BLAST analyses of Cent4 sequences, however, revealed that all 15 had high homology to the poorly characterized LTR retrotransposon RLX\_sela, which was previously shown to be associated with heterochromatic  knobs (Tenaillon et al 2011, Chia et al 2012), suggesting that Cent4 is actually a pericentromeric retrotransposon.

Whole-genome cenH3 chromatin immuno-precipitation data from Wolfgruber (et al 2009) shows no significant over-representation of cent4 compared to five known non-centromeric TE’s, suggesting the cent4 repeat is not involved in kinetichore formation on chromosome 4.  \jri{the jin result is very cool and i wasn't aware of. i moved to discussion though. }

\subsection*{Relatedeness of CentC in the maize genome.}

CentC copies in the maize genome exhibit tremendous diversity: the overall pairwise identity in our alignment was only 65\%, and ~98\% of sites in the alignment had at least 2 variants .  Such diversity led us to ask whether genetic groups of CentC variants could be distinguished. We performed principle coordinate analyses from a genetic distance matrix estimated from our alignment and assigned individual repeats to  genetic clusters following the approach of Patterson et al (2006).  We found 58 significant clusters, but observed no pattern of groupings that revealed chromosome specificity of CentC’s or the impact of historical tetraploidy (Supplemental Table X \jri{add}).

The tandem nature of CentC suggests it increases in copy number through local duplications that produce initially identical copies.  This predicts that local clusters of CentC should be more closely related than CentC from different clusters.  Analysis of genetic and physical distance among CentC repeats on chromosomes 2 and 5 indeed shows this pattern (Figure \ref{heatmap}), revealing significant spatial autocorrelation of CentC variants over distances up to 10 KB (Supplementary Figure SPAGEDI \jri{add}). \jri{need to standardize usage of kb. also i thought autocorrelation was significant up to larger distances?}

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{heatmap.png}
\caption{CentC physical location and genetic relatedness for (a) chromosome 2 and (b) chromosome 5.  On the physical map, red lines show locations of numbered CentC clusters and blue blocks show the location of the active kinetichores.  Scale bar is in Mb.  Below each physical map is shown a heatmap of genetic relatedness of each CentC to (top) other copies within its island of tandem repeats delineated by dotted lines and (bottom) all other copies on the chromosome.  Darker colors indicate higher relatedness.  The total number of CentC in each cluster is shown below the map.  
}
\label{heatmap}    
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage[utf8]{inputenc}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[usenames,dvipsnames]{color}
\usepackage{natbib}
\usepackage{lscape}

@@

Centromere repeats are found in most eukaryotes and play a critical role in kinetochore formation.  Though they exhibit considerable diversity both within and among species, little is understood about the mechanisms that drive centromere repeat evolution.  Here, we use maize as a model to investigate how a complex history involving polyploidy, fractionation, and recent domestication has impacted the diversity of the maize CentC repeat.  We first validate the existence of long tandem arrays of repeats in maize and other taxa in \emph{Zea}.  We show that genetic similarity among CentC copies is highest within these arrays, suggesting that tandem duplications are the primary mechanism for the generation of new copies.  In spite of this, we see little evidence that CentC variants form distinct genetic groups, instead finding that homoplasious mutations have likely homogenized CentC diversity.  We identify Cent4, a repeat initially thought to be specific to the chromosome 4 centromere, as an LTR retrotransposon that has increased drastically in frequency through domestication.  Although the two ancestral subgenomes of maize have contributed nearly equal numbers of centromeres, our analysis shows that a vast majority of all CentC repeats derive from only one of the parental genomes.  Finally, by comparing maize with its wild progenitor teosinte, we find that the abundance of CentC has decreased through domestication.

\keywords{Centromere \and Evolution \and Subgenome \and sweetscience}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section*{Introduction}
\label{intro}
In spite of the rapid growth of sequenced plant genomes, plant centromeres remain poorly understood and relatively cryptic, due largely to their highly repetitive content.  Centromere repeats are highly diverse across taxa and their turnover appears to be very rapid \citep{Melters2012}. However, little is known about the genetic mechanisms that produce centromere repeat diversity.  
Domesticated maize (\emph{Zea mays} ssp. \emph{mays}) has a high quality genome assembly \citep{Schnable2009} including complete sequence of two centromeres \citep{Wolfgruber2009}, and the breadth of research into maize centromeres makes it one of the best systems in which to investigate the processes governing centromere repeat evolution.

Maize centromeres are comprised primarily of the 156bp satellite repeat CentC and a family of retrotransposons (CRM), both of which interact with  kinetochore proteins such as CENH3 \citep{Wolfgruber2009, Zhong2002}, and both repeats show considerable variation in local abundance across taxa \citep{Albert2010}.  But while there is considerable effort in investigating the molecular function of maize centromere repeats \citep{Ananiev1998B, Nagaki2003, Wolfgruber2009}, we know comparatively little about the evolution of the sequences themselves. CRM elements are better understood, including the age and insertion preferences of different CRM families \citep{Wolfgruber2009, Sharma2008}.  In contrast,  studies \citep{Wang2012} have only examined the flanking sequences to CentC islands despite their known association with functional centromeres.  To date, there is no in-depth characterization of the genetic diversity of centromere repeats in the maize genome.

In this paper, we describe the patterns of diversity of centromere repeats across the maize genome.  We investigate the genetic diversity of these repeats, including whether the differential ancestry of maize centromeres \citep{Wang2012} has led to chromosome-specific variation as found in other species \citep{Kawabe2005, Pontes2004} and how genetic relatedness among individual repeats varies spatially along chromosomes and across the genome.  We then compare centromere abundance across a number of maize lines, including to its wild relatives, the teosintes.  We find that CentC copies do not form genetic groups consistent with ancient whole genome duplications or chromosome specificity.  Instead, we show higher genetic similarity within clusters, potentially indicating the predominance of tandem duplications in the formation of new CentC copies.  Lastly, we use low coverage sequencing and cytological analyses data to show that domesticated maize has less CentC than its wild relatives.

@@

We downloaded 218 previously annotated CentC sequences \citep{Ananiev1998B, Nagaki2003} from Genbank.  We then searched the maize genome (5b60, \url{www.maizesequence.org}) with megaBLAST \citep{McGinnis2004} using the 218 annotated CentCs as a reference.  We kept hits with a length of over 140bp and a minimum bit score of 100.  After meeting the bit score threshold, the longest hit was retained.  We defined CentC’s as being in tandem if the CentC’s start location was within 1000bp of the start location of another CentC.

All 12,162 CentC sequences were aligned using 7 iterations of Muscle \citep{Edgar2004} with default parameters.  A Jukes-Cantor distance matrix of all sequences was calculated with PHYLIP (\citep{Felsenstein1989} \url{http://evolution.genetics.washington.edu/phylip.html}), and an unrooted neighbor joining tree was built based on the distance matrix.

We used principle coordinate analysis (PCoA) to cluster CENTC variants based on their genetic distances. Eigenvalues from the PCoA were used to determine the number of statistically significant clusters using the Tracy-Widom distribution \citep{Patterson2006}.

We employed the software SpaGeDi (\citep{Hardy2002}\url{http://ebe.ulb.ac.be/ebe/Software.html}) to estimate the spatial autocorrelation of sequence similarity of CENTC repeats in the completely sequenced centromeres 2 and 5.  We calculated Moran’s I statistic using Jukes-Cantor genetic distance and measures of physical distance between CENTC repeats in base pairs.  Confidence intervals for the values of I were estimated by 20,000 random permutations of the physical distances.

Statistical analyses were performed in R with the packages ape \citep{Paradis2004} and RMTstat \citep{Perry2009}.  We compared clusters to chromosome of origin and syntenic maps of maize ancient tetraploidy \citep{Schnable2011} to determine if the genetic history of maize left a footprint on CentC similarity.

@@

We mapped Illumina reads from a broad panel of \emph{Zea} species \citep{Chia2012,Tenaillon2011}  to a reference consisting of the full complement of 12,162 CentC variants identified in the B73 genome.  We also used previously published whole genome chromatin immunoprecipitation (ChIP) \citep{Wolfgruber2009, Wang2013} with CenH3.  Reads were mapped using  Mosaik v1.0 (\url{https://code.google.com/p/mosaik-aligner/}). We first optimized mapping parameters by relaxing mapping stringency and evaluating the number of successfully mapped reads with each combination.  Consistent with parameters from previous studies mapping reads to repetitive elements \citep{Tenaillon2011}, we required homology to remain at a minimum of 80\%.  For other non-default parameters, we permuted over many values of hash size, alignment candidate threshold,  percent of read aligning, and maximum number of hash positions per seed to find a combination that produced believable alignments.  We selected an optimum combination of parameters just below the parameters where we observed a large increase in the total number of reads aligning (Supplementary Figure \ref{Supp_MPS}).  The parameter boundary at which percentage of reads mapping increased disproportionately was identified, and parameters just below this threshold were selected.  Our final set of parameters for tandem repeats used an initial hash size of 8, an alignment candidate threshold of 15 bases, 20\% percent of mismatching bases, a minimum of 30\% overlap to the reference, and stored the top 100 hits for alignment.  After reads were mapped, we calculated the percentage of total reads hitting the given reference and multiplied this value by the relative genome size of each accession as reported in \citet{Chia2012} and \citet{Tenaillon2011}. The total number of reads mapping did not change drastically when using one random copy of CentC versus the full AGPv2 reference, suggesting that our parameters are sufficiently broad to capture genome-wide CentC.  Because library preparation has an effect on estimates of repeat abundance (see results), we only used individuals from maize HapMap v2 \citep{Chia2012} with libraries prepared using identical methods.

We used a different set of mapping parameters for long repeats such as transposable elements.  Previous studies \citep{Schnable2009} estimated that approximately 85\% of maize genome derives from transposable elements.  Using the short read libraries from \citet{Tenaillon2011}, we selected parameters so that approximately 85\% of the library mapped to the maize transposable element database (\url{maizetedb.org}) with a minimum homology of 80\%.  The final parameters for TEs were a hash size of 10, alignment candidate threshold of 11, 80\% homology excluding non-aligned portions of the read, and a 30\% minimum overlap.

Furthermore, we wanted to know the accuracy of our pipeline in measuring CentC content and therefore designed a simulation to vary CentC content in an artificial genome (code available at: \url{https://github.com/kddistor/dnasims}).  In short, our simulations altered the copy number of CentC repeats over a given genome size, therefore changing the percentage of the genome deriving from the repeat.  Illumina reads were simulated from each of the DNA strings and mapped using our pipeline.  Relative differences in abundance were captured well, though our pipeline underestimated overall quantity of CentC.  Altering the increments of CentC change, we found that our pipeline could accurately capture differences of 0.05\% change in CentC abundance, suggesting that differences greater than 0.05\% are likely to be biologically real (Supplementary Figure \ref{Supp_Accuracy}).

\subsection*{CentC homoplasious mutations simulation}

\jri{needs some basic setup. we start with alignment (?) and then do what to it? what was the code written in? is the code available? if so cite github url}.

We wanted to better understand the diversity present in all CentC copies across the genome.  Looking at our alignment, we found that the sequences had a pairwise percent identify of 65.8 with only 2.2\% of sites identical, suggesting large sequence diversity.  However, these statistics do not inform us whether the amount of diversity observed would be expected due to cumulative mutation over time.  Therefore, we used simulations in python to count the number of sites with shared diversity across copies of CentC (code available on \url{GITHUB} \pb{we should talk about this.}).  Our simulation assumed that CentC has been evolving for the 1 million years since the divergence of maize and \emph{Tripsacum} \citep{Ross-Ibarra2009}, a closely related genus whose centromere repeat shares a large amount of homology \citep{Melters2012}.  We assumed a constant copy number, a mutation rate of $3 x 10^{-8}$, and one generation per year.

@@

Library preparation and sequencing was performed according to the methods cited in Melters et al 2012.  Using those protocols, we sequenced one individual from \emph{mays}, \emph{mexicana}, \emph{parviglumis}, and \emph{Z. luxurians} with Pacific Biosciences (Pacific Biosciences, Menlo Park, CA) technology. Approximately 200Mbp of reads were produced from each cell, and reads with length greater than 600bp were retained for analysis of tandem CentC content using BLAST (Supplementary Table \ref{supp.pacbio}).  CentC copies were considered in tandem if the read had 4 CentC copies within 300bp of each other.

@@

Fluorescent in situ hybridization (FISH) was used to compare abundance of CentC repeats in chromosomes of a maize x \emph{parviglumis} F1 hybrid and a maize x \emph{Z. luxurians} F1 hybrid. FISH protocols closely followed those of X and Y.
\jri{Need Kelly to fill in}

@@

We found a total of 12,162 CentC variants in the maize reference genome and the unassembled BACs.  Of these 12,162 copies, 8,259 were unique over their full length. No CentC occurred more than 10 times in the genome, and the vast majority ($>$75\%, Supplemental Table \ref{supp.occurence}) of non-unique CentC variants occurred only twice.  Of the 2,266 non-unique CentC sequences, only 3 were tandem, identical duplicates.  Genome-wide CentC locations also show that nearly all of the 10,639 CentC copies are found in one of the 248 clusters identified on chromosomes 1-10; only 14 occurred as solo copies.  Cluster size varied from single CENTC copies to 84KB, with a mean of ~7KB (approximately 45 CentC copies). Chromosomes varied greatly in copies of CentC, though we know that centromere assemblies for all of the chromosomes are not complete.   For example, in mapping reads from ChIP with CENH3 to an oat plant with a single maize chromosome (OMA) (\pb{whats the ref here? which paper did jiming send us reads from?}), we find reads from many chromosomes have hits to the unassembled BACs (Supplemental table \ref{gaby}).  In particular, chromosome 6 had many more reads aligning to the BACs than it did to its own centromere, suggesting a particularly lacking assembly.  Examining total repeat number, Chromosome 7 had the most CentC, with 3,200 copies, while chromosome 6 had the fewest with 32 copies.

We used long-read Pacific Biosciences sequencing to verify that most CentC is in tandem arrays. We sequenced whole genome ($\approx 0.1$X) libraries from 4 \emph{Zea} species.  In spite of the low coverage, we recovered reads containing CentC sequence from all four taxa (Supplementary Table \ref{supp.pacbio}).  In one 6.7KB read from the maize reference line B73, for example, we identified approximately 40 independent CentC copies in tandem, and similar arrays were seen in all four \emph{Zea} species analyzed.  These results show that overall structure of the repeats has been maintained for the approximately 140,000 years since the \emph{luxurians}-\emph{mays} divergence \citep{Hanson1996,  Ross-Ibarra2009} and that a majority of CentC is found in tandem arrays (Supplementary Table \ref{supp.pacbio}).

We then identified how many large clusters of CentC were retained from each of the two parental genomes that comprise the extant maize genome, referred to here as subgenome 1 and subgenome 2 (Figure \ref{circos}).  Previous work identified the parental genome for individual chromosomal segments \citep{Schnable2011} and centromeres \citep{Wang2012}.  Because large clusters are less likely to be misassembled, we focused our analyses on the 52 clusters $>$10KB in length (Supplementary Fig Histogram \jri{add}).  While Schnable et al. \citet{Schnable2010} do not assign subgenome status to repetitive regions, we assign clusters to a subgenome if they are flanked by two regions identified as originating from the same  subgenome.  Thirty-eight of these clusters could be assigned to subgenome 1 (out of 43 assignable); of the 17 clusters $>$20KB, 16 were unambiguously assigned to subgenome 1.  Even correcting for the genome-wide overrepresentation of subgenome 1 (62.5\% of assigned base pairs), these results suggest a strong inequality in the origin of large CentC clusters (Fisher's exact test, $p<0.005$ for both 10KB and 20KB clusters).

@@

In addition to CentC, previous studies have described a second high copy number centromere repeat specific to maize chromosome 4 \citep{Page2001}.   BLAST analyses of Cent4 sequences, however, revealed that all 15 had high homology to the poorly characterized LTR retrotransposon RLX\_sela, which was previously shown to be associated with heterochromatic  knobs \citep{Tenaillon2011, Chia2012}, suggesting that Cent4 is actually a pericentromeric retrotransposon.  Whole-genome cenH3 chromatin immuno-precipitation data from  \citet{Wolfgruber2009} shows no significant over-representation of cent4 compared to five known non-centromeric TE’s, suggesting the cent4 repeat is not involved in kinetochore formation on chromosome 4.

\subsection*{Relatedness of CentC in the maize genome.}

CentC copies in the maize genome exhibit tremendous diversity: the overall pairwise identity in our alignment was only 65\%, and ~98\% of sites in the alignment had at least 2 variants .  Such diversity led us to ask whether genetic groups of CentC variants could be distinguished. We performed principle coordinate analyses from a genetic distance matrix estimated from our alignment and assigned individual repeats to  genetic clusters following the approach of \citet{Patterson2006}.  We found 58 significant clusters, but observed no pattern of groupings that revealed chromosome specificity of CentC’s or the impact of historical tetraploidy (Supplemental Table \ref{long}).

The tandem nature of CentC suggests it increases in copy number through local duplications that produce initially identical copies.  This predicts that local clusters of CentC should be more closely related than CentC from different clusters.  Analysis of genetic and physical distance among CentC repeats on chromosomes 2 and 5 indeed shows this pattern (Figure \ref{heatmap}), revealing significant spatial autocorrelation of CentC variants over distances up to 10-50KB (Supplementary Figure \ref{Supp_SpagediChr02} and \ref{Supp_SpagediChr05}).

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{heatmap.png}
\caption{CentC physical location and genetic relatedness for (a) chromosome 2 and (b) chromosome 5.  On the physical map, red lines show locations of numbered CentC clusters and blue blocks show the location of the active kinetichores.  Scale bar is in MB.  Below each physical map is shown a heatmap of genetic relatedness of each CentC to (top) other copies within its island of tandem repeats delineated by dotted lines and (bottom) all other copies on the chromosome.  Darker colors indicate higher relatedness.  The total number of CentC in each cluster is shown below the map.  
}
\label{heatmap}    
\end{figure}

The decreased genetic distance among CentCs in local clusters on chromosome 2 and 5 suggest that many of the genetic groupings discovered in our genome-wide analysis should correspond to local clusters of repeats. We see no evidence of this, however, as repeats within individual clusters are frequently found in different genetic groups as defined by PCoA (Figure \ref{pcoa}).  A comparison of all pairs of CentC reveals a likely explanation: of the $\approx 74$ million possible pairs,  approximately 6 million share $\geq 2$ mutations different from the genome-wide consensus, likely grouping in genetic clusters despite their physical distance.  Looking at several triplets at random from our alignment confirms that two sequences in one PCoA assignment share greater pairwise identity than two sequences adjacent to one another in different PCoA groups.  A simple forward simulation (see \ref{methods} \pb{guessing you want link to github here, or link to section in methods?}) suggests this pattern could be due entirely to homoplasy rather than long-distance movement of CentC repeats.  By stochastically applying mutations to an initially homogeneous group of repeat sequences, we find that plausible parameter values produce $\approx 10$ million pairs of repeats sharing $\geq 2$ mutations.

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{TW_CentCgraph.jpeg}
\caption{Presence of CentC in each of the heirarchical groups.  The 58 clusters found to be statistically significant in forming genetic groups are represented on the x-axis and chromosome of origin for CentC's on the y-axis. The intersection shows the number of CentC's in the group found on that chromosome, scaled logarithmically.  CentC counts from chromosomes whose centromeres were derived from subgenome 1 are colored green and those from subgenome 2 are colored brown.
}
\label{pcoa}    
\end{figure}
